---
title: "A non-linearity-assuming correlation coefficient: XICOR"
author: "Patrick Cherry"
date: "2023/12/30"
categories:
  - code
  - statistics
image: ""
execute:
  freeze: true
---
## Introduction and motivaation


## Load library
```{r setup, message = FALSE}
library(tidyverse)
library(XICOR)
library(psych)
library(nls2)
theme_set(theme_light())
```

## Load data
```{r}
data("pressure")
penguins <- palmerpenguins::penguins_raw
data("peas")
```

### Vapor Pressure of mercury
Data on the relation between temperature in degrees Celsius and vapor pressure of mercury in millimeters (of mercury).
```{r}
glimpse(pressure)
```

```{r}
pressure_point_plot <- pressure %>%
  ggplot(aes(y = pressure, x = temperature)) +
  geom_point() +
  stat_smooth(method = "lm", formula = "y ~ x", color = "orange") +
  stat_smooth(method = "glm", formula = "y ~ I(x^7)", color = "blue") +
  # stat_smooth(method = "nls", formula = "log(y) ~ A - (B / (C + x))", color = "green",
  #             method.args = list(start = c(A = 7, B = 3000, C = 262)), se = FALSE) +
  labs(title = "Plot of non-linar data, vapor pressure of mercury vs. temperature",
       subtitle = "Vapor pressure is know a priori as a non-linear function of temperature.",
       caption = "Linear regression in orange, x^7 glm in blue.")
pressure_point_plot
```

# Non-linear regression using Antoine equation
The following equation comes from the [Antoine equation](https://en.wikipedia.org/wiki/Antoine_equation) that describes the relation between vapor pressure and temperature for pure substances. It's derived from the Clausius–Clapeyron equation, which relates temperature to the derivative of pressure along a substances liquid-vapor coexistence curve.
```{r}
pressure_log_nls <- nls(log10(pressure) ~ A - (B / (C + temperature)), pressure,
    start = c(A = 7, B = 3000, C = 273.15))
summary(pressure_log_nls)
```

```{r}
pressure_point_log_plot <- pressure %>%
  ggplot(aes(y = pressure, x = temperature)) +
  geom_point() +
  stat_smooth(method = "lm", formula = "y ~ x", color = "orange") +
  #stat_smooth(method = "glm", formula = "y ~ I(x^7)", color = "blue") +
  stat_smooth(method = "nls", formula = "y ~ A - (B / (C + x))", color = "green",
              method.args = list(start = c(A = 7, B = 3000, C = 262)), se = FALSE) +
  scale_y_log10() +
  labs(title = "Plot of non-linar data, vapor pressure of mercury vs. temperature",
       y = "pressure, log scale (Torr)",
       x = "temperature (Celsius)",
       subtitle = "Vapor pressure is know a priori as a non-linear function of temperature.",
       caption = "Linear regression in orange, x^7 nls in green")
pressure_point_log_plot
```

```{r}
plot(pressure_log_nls)
```

Let's compare the model fit outputs from these two.
```{r}
pressure_lm <- lm(formula = "pressure ~ temperature", data = pressure)
pressure_glm <- glm(formula = "pressure ~ I(temperature^7)", data = pressure)
```

#### linear model summary
```{r}
summary(pressure_lm); summary(pressure_lm)[["r.squared"]]^0.5
```

#### glm model summary
```{r}
summary(pressure_glm)
```

We can see that the linear summary shows the linear fit is poor: the standard error is high relative to the estimate. Another metric that statistical model practitioners will note is the R-squared, or the correlation coefficient, is low (for data from the physical science): `R^2 = 0.574`. If we square root the R^2, we get a correlation coefficient of `R = 0.758`. But we can see visually from the plot that the data do follow a clear trend.

Meanwhile, the generalized linear model with a `pressure^7` term shows a very low error compared to the estimate. (The error is 34.7 times lower than the estimate.) However, this model summary output lacks a correlation coefficient. __Even though we can get many measures of the fit of a model, if we use anything other than a simple linear model, a correlation coefficient doesn't exist.__ At least a correlation coefficient we are used to: a Pearson r. This is a fact of statistics, and is a consequence of the assumption of Pearson r that the data are linearly related.

### XICOR correlation coefficient

```{r}
xicor(pressure$pressure, pressure$temperature)
```

Ah, but if 

### Palmer Penguins data
```{r}
glimpse(penguins)
```

```{r}
penguin_point_plot <- penguins %>%
  ggplot(aes(y = `Delta 13 C (o/oo)`, x = `Date Egg`)) +
  geom_point() +
  stat_smooth(method = "lm", formula = "y ~ x", color = "orange")
  # stat_smooth(method = "glm", formula = "y ~ x + exp(x)", color = "green",
  #             method.args = list("family" = "inverse.gaussian")) +
  # labs(title = "Plot of non-linar data, vapor pressure of mercury vs. temperature",
  #      subtitle = "Vapor pressure is know a priori as a non-linear function of temperature.",
  #      caption = "Linear regression in orange, x^5 glm in blue.")
penguin_point_plot
```

